\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-22 jasa/paper-jasa.tex
Section 4: further improvements

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{xfrac}
\usepackage{hyperref}
\usepackage{color}

@@

\subsection{Probabilistic classification for approximating ratios}

@@

\glnote{Needs some work.}

@@

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_0^C$, the generalized likelihood ratio test, also known as
the profile likelihood ratio test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\Lambda({\cal D}; \Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
\end{equation}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

Evaluating Eqn.~\ref{eqn:generalized-lr} requires finding the maximum likelihood
estimator
\begin{equation}\label{eq:mle}
    \hat{\theta} = \argmax_\theta p({\cal D} | \theta),
\end{equation}
which is again
made difficult in the likelihood-free setting.  Also, it is not obvious that if
we are working instead with distributions $p(s(\mathbf{x})|\theta)$ that we can
find the same estimators. Fortunately, there is a construction based on $s$ that
works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is the same as the
value that maximizes the likelihood ratio with respect to $p({\cal D}|\theta_1)$,
for some fixed value of $\theta_1$. This allows us to use
Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
estimate as
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized}
transformation $s$ of $X$ in terms of $(\theta,\theta_1)$ that is monotonic
with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
cancels Jacobian factors that vary with $\theta$.

@@

\glnote{Needs to be clarified.}
\glnote{Do we put back the algorithm snippet? I think this could be directly
explained in words.}

In order to provide parameter inference in the likelihood-free setting, we must
train a family of classifiers parametrized by $\theta_0$ and $\theta_1$, the
parameters associated to the null and alternate hypotheses, respectively. While
this could be done independently for all $\theta_0$ and $\theta_1$, it is
desirable and convenient to have a smooth evolution of the classification score
as a function of the parameters. Thus, we anticipate a single learning stage
based on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$.
Somewhat unusually, the unknown values of the parameters are taken as input to
the classifier; their values will be specified via the enveloping (generalized)
likelihood ratio of Eqn.~\ref{eqn:generalized-lr}. We denote the learned family of
classifiers $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$, and anticipate the training based
roughly on the following algorithmic flow.

% \begin{algorithm}[ht]
% \caption{Training of the parametrized classifier.}\label{alg:training}
% \begin{algorithmic}
% \STATE initialize trainingData to \{\}
% \FOR{ $\theta_0$ in $\Theta$ }
% 	\FOR{ $\theta_1$ in $\Theta$ }
% 		\STATE generate $x_i^0 \sim p(x|\theta_0)$
% 		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
% 		\STATE generate $x_i^1 \sim p(x|\theta_1)$
% 		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
% 	\ENDFOR
% \ENDFOR
% \STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
% \end{algorithmic}
% \end{algorithm}%\vspace{-1.5em}

While the function $p(\mathbf{x}|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
the expected squared loss based on training data produced according to
Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0
\ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world
classifier with finite capacity. This is left as an area for future work.

@@

\glnote{Needs to be clarified as well.}

Once the classifier is trained, we can use the generative model together with a
univariate density estimation technique (e.g. histograms or kernel density
estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter
points. For an single parameter point, this is a tractable univariate density
estimation problem. The challenge comes from the need to estimate this density
for all values of $\theta$. A straight forward approach would be to run the
generative model on demand for any particular value of $\theta$. In the context
of a likelihood fit this would mean that the optimization algorithm that is
trying to maximize the likelihood with respect to $\theta$ needs access to the
generative model $p(x|\theta)$. This can be  impractical when the generative
model is computationally expensive or has high-latency (for instance some human
intervention is required to reconfigure the generative model).  In  HEP with a
fixed classifier, it has become common  to interpolate the distribution between
discrete values of $\theta$ in order to produce a continuous parametrization for
$\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.

@@

One can easily imagine a number of approaches to embedding the classifier and
estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those
approaches will depend critically on the dimensionality of $\theta$ and the
computational cost of the generative model. We leave a more general strategy for
this overarching optimization problem as an area of future work.

@@

In high energy physics, we are often searching for some class of events,
generically referred to as \textit{signal}, in the presence of a separate class
of \textit{background} events.  For each event we measure some quantities $x$
that have corresponding distributions $p_b(x|\nu)$ for background and
$p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
uncertainties in the underlying physics prediction or response of the
measurement device. The total model is a mixture of the signal and background,
and $\mu$ is the mixture coefficient associate to the signal component.
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where
the null corresponds to $\mu=0$, and the generalized likelihood ratio (profiling
over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test
statistic~\citep{Cowan:2010js}. This approach was used for the discovery of the
Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.

@@

In this section we generalize the capacity focusing technique of training
classifiers to discriminate between components of a mixture model. First, we
generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
\begin{equation}
p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
\end{equation}
It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems.
\begin{eqnarray}\label{eq:decompose}
\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}
\\
%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\ \label{eq:decomposedResult}
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1}
\end{eqnarray}
The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.  The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio). Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{xfrac}
\usepackage{hyperref}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\subsection{Probabilistic classification for approximating ratios}
\label{sec:clf-for-ratios}

@@

@@

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_1$ (possibly such that $\Theta_0 \subseteq \Theta_1$), the
generalized likelihood ratio test, also known as the profile likelihood ratio
test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\Lambda({\cal D}; \Theta_0; \Theta_1) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta_1} p({\cal D} | \theta)} \; .
\end{equation}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

Evaluating the generalized likelihood ratio as defined by
Eqn.~\ref{eqn:generalized-lr} requires finding for both the numerator and the
denominator the maximum likelihood estimator
\begin{equation}\label{eq:mle}
    \hat{\theta} = \argmax_\theta p({\cal D} | \theta).
\end{equation}
Again, this is made difficult in the likelihood-free setting and it is not
obvious that we can find the same estimators if we are working instead with
$p(s(\mathbf{x})|\theta)$. Fortunately, there is a construction
based on $s$ that works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is
the same as the value that maximizes the likelihood ratio with respect to
$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$. This allows us to
use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
estimate as
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized}
transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
cancels Jacobian factors that vary with $\theta$.

@@

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parametrized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:clf-for-ratios}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.

\begin{algorithm}[t]
\caption{Training of a parametrized classifier.}\label{alg:training}
    \begin{algorithmic}
        \State ${\cal L := \{ \}}$
        \For{ $\theta_0$ in $\Theta_0$ }
        	\For{ $\theta_1$ in $\Theta_1$ }
        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$
        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=0) \}$ to ${\cal L}$
        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$
        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=1) \}$ to ${\cal L}$
        	\EndFor
        \EndFor
        \State Learn $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$
    \end{algorithmic}
\end{algorithm}

While the function
$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
minimize the expected squared error loss based on training data produced
according to Algorithm~\ref{alg:training}, it is not clear how training data
from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
world classifier with finite capacity. This is left as an area for future work.

\glnote{Shall we keep Algorithm~\ref{alg:training} or rather give a sketch in
words? In its current state, the algorithm does not explain how to build ${\cal
L}$ when the parameter spaces are continuous. It also does not specify how many
samples should be generated per parameter value.}

@@

Once the classifier is trained, we can use the generative model together with a
univariate density estimation technique (e.g. histograms or kernel density
estimation) to approximate $p(s|\theta)$ for specific parameter
points. For a single parameter point, this is a tractable univariate density
estimation problem. The challenge comes from the need to calibrate this density
for all values of $\theta$. A straight forward approach would be to run the
generative model on demand for any particular value of $\theta$. In the context
of a likelihood fit this would mean that the optimization algorithm that is
trying to maximize the likelihood with respect to $\theta$ needs access to the
generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
model is computationally expensive or has high-latency (for instance some human
intervention is required to reconfigure the generative model).  In HEP, with a
fixed classifier, it has become common  to interpolate the distribution between
discrete values of $\theta$ in order to produce a continuous parametrization for
$p(s | \theta)$~\citep{Cranmer:2012sba}.
One can easily imagine a number of approaches to embedding the classifier and
estimating the density $p(s|\theta)$ and the relative merits of those
approaches will depend critically on the dimensionality of $\theta$ and the
computational cost of the generative model. We leave a more general strategy for
this overarching optimization problem as an area of future work.

\glnote{Sub-sections 4.1 and 4.2 could directly be inlined within section 4. What do you think?}

@@

@@

In high energy physics, we are often searching for some class of events,
generically referred to as \textit{signal}, in the presence of a separate class
of \textit{background} events.  For each event we measure some quantities $x$
that have corresponding distributions $p_b(x|\nu)$ for background and
$p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
uncertainties in the underlying physics prediction or response of the
measurement device. The total model is a mixture of the signal and background,
and $\mu$ is the mixture coefficient associate to the signal component.
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where
the null corresponds to $\mu=0$, and the generalized likelihood ratio is used as a test
statistic.

@@

In this section we generalize the capacity focusing technique of training
classifiers to discriminate between components of a mixture model. First, we
generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
\begin{equation}
p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
\end{equation}
It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems.
\begin{align}
\frac{p(x|\theta_0)}{p(x|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} \label{eq:decomposedResult}
\end{align}
The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.  The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio). Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
